---
title: "Significant Predictors"
subtitle: "October 25"
author: "Cindy Ha, Willie Xie, Erica Zhang"
format: pdf
editor: visual
execute: 
  echo: false
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(skimr)
library(lubridate)

data = read_csv('data/raw_data/owid-covid-data.csv')

g20 = c('Argentina', 'Australia', 'Canada', 'China', 'France', 'Germany',
        'India', 'Italy', 'Japan', 'South Korea', 'Mexico', 'Russia',
        'Saudi Arabia', 'South Africa', 'Turkey', 'United Kingdom',
        'United States')
g24 = c('Argentina', 'China', 'Colombia', 'Ecuador', 'Ethiopia', 'India',
        'Mexico', 'Morocco', 'Pakistan', 'Philippines', 'South Africa',
        'Sri Lanka')
data_cur = data %>%
  filter(location %in% c(g20, g24)) %>%
  mutate(G20 = location %in% g20, G24 = location %in% g24)
```

# Initial Dataset Information

**The number of observations and variables in the initial dataset.**

```{r}
dim(data_cur)
```

# Remove nonvalue predictors

**Only iso_code and tests_units hold no significant value.**

```{r}
y = data_cur %>% relocate(iso_code, tests_units, .before = continent)
head(y, n = 5)

x = data_cur %>% select(-c(iso_code, tests_units))
```

# Remove nonsignificant predictors

**The predictors with large missingness.**

```{r}
results = data_cur %>% skim_without_charts()
names_filt = results$skim_variable[results$complete_rate < 0.7]

names_filt
```

**The predictors with large missingness, but are significant to predicting `new_cases` and thus, kept.**

```{r}
y = x %>% select(new_cases, any_of(names_filt))

cor(y, use = "pairwise.complete.obs") %>% round(digits = 3)

c('total_tests', 'new_tests',
  'positive_rate', 'total_vaccinations')

names_filtn = setdiff(names_filt,
                      c('total_tests', 'new_tests',
                        'positive_rate', 'total_vaccinations'))

x = data_cur %>% select(-c(iso_code, tests_units)) %>%
  select(-any_of(names_filtn))
```

## Number of observations with hosp_patients missingness by country

```{r}
sort(table(data_cur$location[is.na(data_cur$hosp_patients)]))
```

## Number of observations with icu_patients missingness by country

```{r}
sort(table(data_cur$location[is.na(data_cur$icu_patients)]))
```

# Remove multicollinearity predictors

```{r}
cor(x %>% select(-c(continent, location, date, G20, G24)),
    use = "pairwise.complete.obs") %>%
  round(digits = 3)
```

**The predictors with multicollinearity issues.**

```{r}
names_col = c('iso_code', 'median_age', 'aged_65_older', 'aged_70_older',
              'human_development_index', 'new_cases_smoothed', 'new_deaths_smoothed',
              'new_deaths_smoothed_per_million', 'new_cases_smoothed_per_million')

names_col

x = x %>% select(-any_of(names_col))
```

# Final dataset

**The number of observations and variables in the final dataset.**

```{r}
dim(x)

colnames(x)
```
